# Introduction {#intro}


**What factors do you consider when making a decision?**

```{r, echo = F}
#knitr::include_url("https://docs.google.com/spreadsheets/d/e/1RxYCQaUkklkHyiAMPvktNsR5QhG7XhFtOBHZGn2-RIU/pubhtml?gid=665289068&amp")
```

<iframe src="https://docs.google.com/spreadsheets/d/1RxYCQaUkklkHyiAMPvktNsR5QhG7XhFtOBHZGn2-RIU/htmlembed/sheet?gid=665289068&range=B2:20" width="700" height="895" frameborder="0" marginheight="0" marginwidth="0"> </iframe>




You can label chapter and section titles using `{#label}` after them, e.g., we can reference Chapter \@ref(intro). If you do not manually label them, there will be automatic labels anyway, e.g., Chapter \@ref(methods).

Figures and tables with captions will be placed in `figure` and `table` environments, respectively.

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

Reference a figure by its code chunk label with the `fig:` prefix, e.g., see Figure \@ref(fig:nice-fig). Similarly, you can reference tables generated from `knitr::kable()`, e.g., see Table \@ref(tab:nice-tab).

```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(iris, 20), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```

You can write citations, too. For example, we are using the **bookdown** package [@R-bookdown] in this sample book, which was built on top of R Markdown and **knitr** [@xie2015].
